---
title: "BIOL4800/6220_Fall2022_RTutorial"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

# R Basic

## Introduction

### R as a calculator

```{r}
1 + 100
```
```{r}
2^2
```
```{r}
7/2
```
```{r}
4-1
```
```{r}
2*2
```
```{r}
3+5*2
```
```{r}
(3+5)*2
```
```{r}
2e2
```
```{r}
2/100000
```
```{r}
sin(1)
```
```{r}
log(1)
```
```{r}
exp(0)
```
```{r}
1 == 1
```
```{r}
1 >=1
```
```{r}
1 < 1
```
```{r}
1 != 2
```
```{r}
x <- 10
x
```
```{r}
x + x
```
```{r}
x -5
```
```{r}
x^2
```
```{r}
y <- x + x
y
```
### Vectorization

```{r}
1:5
```
```{r}
2^(1:5)
```
```{r}
v <-1:5
v
```
```{r}
log(v)
```

### Environment

```{r}
ls()
```
```{r}
rm(x)
```
```{r}
ls()
```

## Package Management
```{r}
installed.packages()
```
```{r}
#install.packages("vegan", dependencies=TRUE)
```
```{r}
#library(vegan)
```
```{r}
#remove.packages("vegan")
```
```{r}
#library(vegan)
```

## Project Management

1. Set up management structure (i.e. "data", "src", "results", "doc")
2. Discuss avoiding redundancy of files
3. Make sure to use version control (i.e. Git)

## Getting Help
```{r}
?plot
```

```{r}
help(plot)
```

```{r}
?"<-"
```

```{r}
vignette("FAQ-vegan")
```

```{r}
citation("vegan")
```
## Data and their formats

```{r}
cats <- data.frame(coat = c("calico","black","tabby"),
                   weight = c(2.1,5,3.2),
                   likes_string = c(1,0,1))
```
```{r}
cats
```
```{r}
getwd()
```
```{r}
#write.csv(cats, file = "./data/cats.csv")
```
```{r}
#read.csv("./data/cats.csv")
```
```{r}
cats$coat
```
```{r}
cats$weight
```
```{r}
cats$weight*10
```
```{r}
log(cats$weight)
```
```{r}
logweight = log(cats$weight)
cbind(cats,logweight) #adds column to the cats table
```
```{r}
cats <- cbind(cats,logweight)
```
```{r}
cats
```
```{r}
paste("My cat is " , cats$coat, " and it weighs " , cats$weight, "kg. ", sep="")
```
### Data Types

```{r}
typeof(cats$coat)
```
```{r}
typeof(cats$weight) #double = number with decimal places
```
```{r}
typeof(cats$likes_string)
```
```{r}
typeof(3.14159)
```
```{r}
typeof(1i)
```
```{r}
typeof(FALSE)
```
```{r}
typeof(TRUE)
```
```{r}
typeof(T)
typeof(F)
```
```{r}
typeof(cats)
```
```{r}
class(cats)
```
```{r}
#file.show("./data/cats.csv")
```

```{r}
cats$likes_string <- as.logical(cats$likes_string)
typeof(cats$likes_string)
```
```{r}
cats$likes_string
```
```{r}
c("a","b")
```
```{r}
ab <- c("a","b")
ab
```
```{r}
c(ab, "c")
```
```{r}
c(ab, ab)
```
```{r}
1:10
```
```{r}
seq(10)
```
```{r}
z <- seq(10)
head(z, n=3)
```
```{r}
length(z)
```
```{r}
class(z)
```
```{r}
typeof(z)
```
```{r}
seq(0,100, by=5)
```
### Data Frames
```{r}
cats
```
```{r}
str(cats$coat)
```
```{r}
coats <- c("tabby","tortoiseshell" , "tortoiseshell", "black", "tabby")
coats
```
```{r}
str(cats$coat)
```

```{r}
factor(coats)
```
```{r}
categories <- factor(coats)
class(coats)
class(categories)
```
```{r}
str(categories)
```

### Lists

```{r}
list_example <- list(title="numbers", numbers =1:10, data=TRUE)
```
```{r}
list_example
```
```{r}
another_list <- list(1,"a",TRUE,1+1i)
```
```{r}
another_list
```

```{r}
typeof(list_example)
```
```{r}
class(list_example)
```
```{r}
data.frame(list_example)
```
```{r}
cats
```
```{r}
cats[,3] #shows the 3rd column
```
```{r}
cats[,2]
```
```{r}
cats[3,] #shows the 3rd row
```
```{r}
cats[2:3,c(1,3)] #rows 2 and 3, columns 1 and 3
```

### Matrices

```{r}
#matrices are for math
matrix_example <- matrix(0, ncol=5, nrow=3)
```
```{r}
matrix_example
```
```{r}
class(matrix_example)
```
```{r}
typeof(matrix_example)
```
```{r}
str(matrix_example)
```
```{r}
dim(matrix_example) #output is row column
```
```{r}
ncol(matrix_example)
nrow(matrix_example)
```
```{r}
class(data.frame(matrix_example))
```
```{r}
df_example <- data.frame(matrix_example)
```
```{r}
df_example
```

### Subsetting

```{r}
p <- c(2.3,6.9,4.0,23,1)
p
```
```{r}
names(p) <- c("a", "b", "c", "d", "e")
p
```
```{r}
p[1] #in subsetting p[] gives you entry
p[2:4]
```
```{r}
p[c(1,5)]
```
```{r}
p[c(1,1,1,3,5,5)]
```
```{r}
p[c(6)]
```
```{r}
p[-3] #everything except 3
```
```{r}
p[c(-1,-5)]
```
```{r}
p[-(2:4)]
```
```{r}
p[c("a", "c")]
```
```{r}
p[c(T,F,T,F,T)] #1,3,5
```
```{r}
p[names(p) != "c"] #where names do not equal(!=) c
```

### Factors

```{r}
f <- factor(c("a", "b", "c", "d", "e"))
f
```
```{r}
f[f == "a"] #== is equal to
```
```{r}
f[1:3]
```
```{r}
f[f %in% c("b", "c")] #if f contains b or c return it
```
```{r}
f[-3]
```
```{r}
f2 <- factor(c("a","a","b","c","c"))
f2
```
```{r}
f2[f2 == "a"]
```
```{r}
f2[f2 %in% c('a','c')] # f2 %in% searching in f2
```

### Matrices Revisited

```{r}
set.seed(65) #makes matrix reproduceably random
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m
```
```{r}
m[3:4, c(3,1)]
```
```{r}
m[,c(3,4)]
```

### Lists Revisted

```{r}
xlist <- list(a="BIOL48006220", b=seq(1,10, by = 0.5), data="Grades")
xlist
```
```{r}
xlist[1:2]
```
```{r}
xlist['a']
```
```{r}
xlist[['a']] #only values of a
```
```{r}
xlist['b']
xlist[['b']]
```
```{r}
xlist$b #name$ returns only the values
```

### Data Frames Revisited

```{r}
gp <- read.csv('~/OneDrive - East Carolina University/biol48006200_Fall2022/data/forest_area_sq_km.csv') #use tab to find path
head(gp, n=10L) #10 lines L=lines
```
```{r}
head(gp[3], n=10L)
```
```{r}
nrow(gp)
```
```{r}
ncol(gp)
```
```{r}
head(gp[["country"]], 10L) #head=header
```
```{r}
gp$X1998 #$ pulls up all headers
```
```{r}
gp[1:3, 2:5] #rows
```
```{r}
gp[c(1:3), c(2:5)] #columns
```
```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100),] #returns countries that had forest area of less than 100 in 1989 and more than 100 in 2019 (messed up from the millions)
```

### Conditionals and Flow
```{r}
n <- 10

if (n < 10) {
  print("n is less than 10")
} else if (n > 10) {
  print("n is greater than 10")
} else {
  print("n is equal to 10")
}
```
```{r}
for (n in seq(0,20)) {
  if (n < 10) {
    print("n is less than 10")
  } else if (n > 10) {
   print("n is greater than 10")
  } else {
    print("n is equal to 10")
  }
} #for loops in R with conditionals
```
```{r}
g <- 0
while (g <= 10) {
  print(paste(g,"is less than or equal to 10"))
  g <- g+1
} #While statement stops when g is no longer less than or equal to 10 (only returns WHILE g is less than or equal to 10)
```

## Plotting and Figures

```{r}
#install.packages("ggplot2")
```
```{r}
library(ggplot2)
```
```{r}
#gp <- read.csv('~/OneDrive - East Carolina University/biol48006200_Fall2022/data/gapminder/gapminder_all.csv')
head(gp)
```
```{r}
#ggplot(data=forestdata, mapping=aes(x=X1989, y=X2019)) + geom_point()
```


```{r}
library(ggplot2)
```
```{r}
#gp <- read.csv('~/OneDrive - East Carolina University/biol48006200_Fall2022/data/gapminder/gapminder_all.csv')
head(gp)
```
```{r}
#library(ggplot2)
#ggplot(data=gp, mapping=aes(x=gdpPercap_1952, y=pop_1952)) + geom_point()
```
```{r}
#ggplot(data=gp, mapping=aes(x=gdpPercap_2002, y=pop_2002)) + geom_point()
```
```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")
head(gapminder)
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) +geom_line(mapping=aes(color=continent)) + geom_point()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point() + scale_x_log10()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.5) + scale_x_log10() #alpha is transparency-can see how dense the plot is
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(alpha=0.25, color="purple", size=1.5, pch=9) + scale_x_log10() + geom_smooth(method=lm, color="gold", size=1.25) #pch=shape
```
```{r}
africas <- gapminder[gapminder$continent== "Africa", ] #square brackets=indexing/subsetting
head(africas)
```
```{r}
ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line() +
  facet_wrap( ~ country)
```
```{r}
ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line(color="red") +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(
    x = "Year",
    y = "Life Expectency",
    title = "Life Expectency Over Time in African Countries"
  )
```
```{r}
AfricanLifeExp <- ggplot(data=africas, mapping=aes(x=year, y=lifeExp)) +
  geom_line(color="red") +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(
    x = "Year",
    y = "Life Expectency",
    title = "Life Expectency Over Time in African Countries"
  )
```

```{r}
ggsave(filename= "data/AfricanLifeExp.png", plot=AfricanLifeExp, width=24, height=40, dpi=300, units="cm")
```

```{r}
pdf(file= "data/AfricanLifeExp.pdf", width=24, height=40)
plot(AfricanLifeExp)
dev.off()
```

```{r}
write.table(gapminder, file="data/gapminder_web.csv", sep=",")
```

```{r}
write.csv(africas, file="data/gapminder_web_africas.csv")
```

### Fancy Plots
```{r}
#install.packages(c("ggridges", "viridis", "hrbrthemes"), dependencies = T)
```
```{r}
library(ggridges)
library(ggplot2)
library(viridis)
library(hrbrthemes)

#ggplot(lincoln_weather, aes(x='Mean Temperature [F]', y='Month', fill=..x..)) +
 # geom_density_ridges_gradient(scale=3, rel_min_height=0.01) +
  #scale_fill_viridis(name= 'Temp [F]', option="C") +
  
```

```{r}
#install.packages(c("ggstatsplot", "palmerpenguins", "tidyverse"), dependencies = T)
library(ggstatsplot)
library(tidyverse)
```

```{r}
gapminder2 <- drop_na(gapminder)
gapminder2
```
```{r}
#install.packages("vioplot", dependencies = T)
```

```{r}
library(vioplot)
with(gapminder2, vioplot(
  lifeExp~continent, col="blue"
))
```
## Statistics

### Additional ways of importing/reading data

```{r}
rand <- c(12, 54, 98, 65, 38)
rand
```
```{r}
sum(rand)
length(rand)
avg.rand <- sum(rand)/length(rand)
avg.rand
```
```{r}
mean(rand)
```
```{r}
sort(rand)
```
```{r}
min(rand)
max(rand)
```
```{r}
cumsum(rand)
```
```{r}
diff(rand)
```
```{r}
rand[2]
rand[3:4]
```
```{r}
rand*2
```

### Analyses

#### Univariate Statistics

1. Categorical Data
  a. Barplots

```{r}
beer <- c(3,4,1,1,3,4,3,3,1,3,2,1,2,1,2,3,2,3,1,1,1,1,4,3,1)
length(beer)
```

```{r}
table(beer)
```

```{r}
#barplot((table(beer), xlab="Beer", ylab="Frequency")
```

```{r}
barplot(table(beer)/length(beer), xlab="Beer", ylab="Proportion") #proportions insteaad of frequencies
```

  b. Pie Charts
```{r}
pie(table(beer), main="Beer preference by students")
```
```{r}
pedes <- c(1,2,3,52,2,4,35,45,23,34,6,4,64,23,12,14,17,34)
```

2. Numerical Data
  a. Stem-and-leaf plots
```{r}
stem(pedes)
```
  b. Strip Chart
```{r}
#stripchart(pedes, method="stacks")
```
3. Measures of Center
  a. Mean
```{r}
mean(pedes)
```
```{r}
mean(gapminder$lifeExp)
```

  b. Median
```{r}
median(pedes)
```
```{r}
median(gapminder$lifeExp)
```
```{r}
with(gapminder, median(lifeExp))
```

  c. Mode
```{r}
which(table(pedes) == max(table(pedes)))
```

4. Variation
  a. Range
```{r}
range(pedes)
```
```{r}
diff(range(pedes)) #summary of range
```
  b. Variance
```{r}
var(pedes)
```
```{r}
sd(pedes)
```
  c. IQRs interquartile range
```{r}
IQR(pedes)
```
  d. Z-scores
```{r}
scale(pedes)
```
  e. Summary
```{r}
summary(pedes)
```
5. Plot
  a. Histograms
```{r}
hist(pedes, breaks="scott") #scott finds optimal number of bins
```
```{r}
hist(pedes, breaks="scott", prob=TRUE)
lines(density(pedes)) #shows probabilites instead of proportions
```
  b. Box Plots
```{r}
boxplot(pedes)
```
```{r}
summary(boxplot(pedes))
```

#### Bivariate and Multivariate Statistics
1. Plotting and Regression
  a. Box Plotting
```{r}
spid.gen <- read.csv('~/OneDrive - East Carolina University/biol48006200_Fall2022/data/spider_genitalia (1).csv', header=T)
spid.gen
```
```{r}
boxplot(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat)
```

  b. Scatter Plotting
```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
```
  c. 
```{r}
spid.reg  <- lm(spid.gen$left.bulb~spid.gen$right.bulb)
```
  
```{r}
summary(spid.reg)
```

```{r}
#plot(spid.gen$left.bulb ~ spid.gen$right.bulb, pch=as.character(spid.gen$habitat))
#abline(spid.reg)
```
  d. Correlation Coefficients and Spearman Rank Correlation
```{r}
cor.gen <- with(spid.gen, cor(left.bulb, right.bulb))
cor.gen^2 #correlation coefficient, r-squared value
```
```{r}
spearman.cor.gen <- with(spid.gen, cor(left.bulb, right.bulb, method="spearman"))
spearman.cor.gen^2 #correlation coefficient, r-squared value
```
  e. Residuals
```{r}
residuals(spid.reg)
```

  f. Transformations
```{r}
plot(log10(spid.gen$left.bulb)~spid.gen$right.bulb)
# log.left.ball <- log10(spid.gen$left.bulb)
```
2. Comparing Discrete Treamtnet Effects
  a. Chi-Squared
```{r}
obs_weighted <- c(4, 15, 6, 15, 18, 2)
obs_fair <- c(10, 10, 10, 10, 10, 10)
exp <- c(.16, .17, .16, .17, .17, .17)
chisq.test(obs_weighted, p=exp)
chisq.test(obs_fair, p=exp)
```
  b. T-tests
```{r}
#H0 - true mean is equal to 0, but you can set true mean to other values with mu=
t.test(spid.gen$carapace.length)

mean(spid.gen$carapace.length)

t.test(spid.gen$carapace.length, mu=29)
```
```{r}
#two sample t-test
habitat.t2 <- t.test(spid.gen$left.bulb ~ spid.gen$habitat)
habitat.t2
```
```{r}
habitat.tless <- t.test(spid.gen$left.bulb, alternative = "less")
habitat.tless
```
```{r}
habitat.tgreater <- t.test(spid.gen$left.bulb, alternative = "greater")
habitat.tgreater
```

  c. Analysis of Variance (ANOVA)
```{r}
gen.lm <- lm(spid.gen$left.bulb ~ spid.gen$habitat)
summary(gen.lm)
```
```{r}
anova(gen.lm)
```
```{r}
gen.anova <- aov(spid.gen$left.bulb ~ spid.gen$habitat)
summary(gen.anova)
```

```{r}
gapminder
```

```{r}
gapminder.lifeexp.continent.anova <- aov(gapminder$lifeExp ~ gapminder$continent)
summary(gapminder.lifeexp.continent.anova)
```
```{r}
gapminder.tukey.two.way <- TukeyHSD(gapminder.lifeexp.continent.anova)
gapminder.tukey.two.way
```
  d. Analysis of Covariance (ANCOVA)
```{r}
boxplot(spid.gen$left.bulb~spid.gen$habitat)
```
```{r}
gen.ancova <- lm(spid.gen$left.bulb ~ spid.gen$habitat*spid.gen$carapace.length)
summary(gen.ancova)
```
```{r}
gapminder.lifeexp.continent_GDP.ancova <- aov(gapminder$lifeExp ~ gapminder$continent + gapminder$gdpPercap)
summary(gapminder.lifeexp.continent_GDP.ancova)
```
```{r}
#install.packages("multcomp", dependencies = T)
library(multcomp)
#posthoc.gapminder.ancova <- glht(gapminder.lifeexp.continent_GDP.ancova, linfct = mcp(technique="Tukey"))
```

3. Principle Component Analysis

```{r}
sp.matrix <- with(spid.gen, cbind(left.bulb, right.bulb, carapace.length, leg4.length))
sp.matrix
```
```{r}
sp.pca <- princomp(sp.matrix, cor=TRUE)
summary(sp.pca)
```
```{r}
loadings(sp.pca)
```
```{r}
biplot(sp.pca)
```
```{r}
sp.pca$scores
```

```{r}
#install.packages("ggfortify", dependencies = T)
library(ggfortify)
sp.pca.plot <- autoplot(sp.pca,
                        data = spid.gen,
                        colour ='habitat')
sp.pca.plot
```













